---
title: "Tiger Milkweed Butterflies (genus Danaus) Statistical Project"
author: "Julia Zheng & Kate Skocelas"
date: "10/15/2020"
output:  
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Read in Raw Data
```{r input files}
#setwd("C:\\Dropbox\\Classes\\Classes\\IBIO 830 stats\\project1")
setwd("~/Desktop/stats/Project1/IBIO830---Statistical-Project")

# load raw data from csv file
butterfly_raw <- read.csv("Danaus.csv", header = TRUE)
```

### Load required libraries
```{r load libraries, message=FALSE}
if (!require(RColorBrewer)) install.packages('RColorBrewer')
library(RColorBrewer)

if (!require(rworldmap)) install.packages('rworldmap')
library(rworldmap)

if (!require(scales)) install.packages('scales')
library(scales)

if (!require(dplyr)) install.packages('dplyr')
library(dplyr)

if (!require(ggplot2)) install.packages('ggplot2')
library(ggplot2)
```

## Data Cleaning and Exploration
```{r clean data, echo=TRUE}
# Rename column where there is a typo 
names(butterfly_raw)[names(butterfly_raw) == "phyenotype..metamorphosis.stage."] <- "metamorphosis_stage"

# check for NAs in the data
sum(is.na(butterfly_raw)) # 132 NAs found 

# Remove NAs in the data columns we are looking at
butterfly_data <- butterfly_raw
butterfly_data <- butterfly_data[!is.na(butterfly_data$latitude),]
butterfly_data <- butterfly_data[!is.na(butterfly_data$longitude),]
butterfly_data <- butterfly_data[!is.na(butterfly_data$positional_accuracy),]
butterfly_data <- butterfly_data[!is.na(butterfly_data$scientific_name),]
butterfly_data <- butterfly_data[!is.na(butterfly_data$metamorphosis_stage),]
butterfly_data <- butterfly_data[!is.na(butterfly_data$observed_on),]
```

### Initial plotting
```{r look at data}
# look at the data
head(butterfly_raw)
```

#### Metamorphic stage of samples
```{r count stage samples}
# number of samples in each metamorphic stage
count(butterfly_data, butterfly_data$metamorphosis_stage)
```

```{r plot metamorphic stage}
# Create bar plot of metamorphic stage
ggplot(butterfly_data, aes(x=metamorphosis_stage))+
  geom_bar(col="black", fill= c("turquoise2", "darkorchid4", "orange2", "red"))+
  labs(x="Metamorphic Stage", y="Number of Samples", title="Metamorphic Stage of Samples")+
  theme_classic()
```

There are significantly more samples of organisms in the "butterfly" metamorphic stage than in any of the other metamorphic stages. This may represent a sampling bias in the dataset. Individuals in the "butterfly" stage may be easier for amateur naturalists to spot, or they may be more appealing to photograph and log. In the statistical tests section, we use a chi-squared goodness of fit test to test the null hypothesis that our data set does not accurately fit the expected average lifecyle durations of genus Danaus butterflies.


#### Metamorphic stage of sample compared to observation date
```{r metamorphic stage vs year observed}
# add year of observation to data frame
butterfly_data$year <- substr(butterfly_data$observed_on, nchar(butterfly_data$observed_on)-3 ,nchar(butterfly_data$observed_on))

# get the number of samples collected in each year
count(butterfly_data, butterfly_data$year)

#plot stage against year of observation
ggplot(butterfly_data, aes(x=year, fill=metamorphosis_stage))+
  geom_bar()+
  labs(x="Year", y="Number of Samples", title="Metamorphic Stage of Samples by Year Observed")+
  theme_classic()
```
The distribution of samples by year appears to follow the pattern expected by the rise in popularity of the iNaturalist website. We suspect that 2020 has fewer samples (so far) than 2019 because the year is not yet complete.

```{r metamorphic stage vs month observed}
# add month of observation to data frame
butterfly_data$month <- substr(butterfly_data$observed_on, 1,2)
for(i in 1:nrow(butterfly_data)) {
  month <- butterfly_data$month[i]
  last_letter <- substr(month, 2,2)
  if(last_letter == "/") {
    month <- substr(month, 1,1)
  }
  butterfly_data$month[i] <- month
}

# get the number of samples collected in each month
butterfly_data$month_fac <- factor(butterfly_data$month, levels=c("1","2","3","4","5","6","7","8","9","10","11","12"))
count(butterfly_data, butterfly_data$month_fac)

#plot stage against month of observation
ggplot(butterfly_data, aes(x=month_fac, fill=metamorphosis_stage))+
  geom_bar()+
  labs(x="Month", y="Number of Samples", title="Metamorphic Stage of Samples by Month Observed")+
  theme_classic()
```
The observations appear to have a bimodal distribution. The first peak is in March, and the second peak is in August. If these two normal distributions exist, they may indicate the peak butterfly seasons in two distinct climates. In the statistical tests, we test the hypothesis that a sample's observation month is related to the latitude of the observation location, because butterflies are most active in warm weather.

#### Positional accuracy vs. latitude and longitude
Positional accuracy in iNaturalist describes how far away the submitted observations are in meters from the recorded longitude and latitude coordinates. Therefore, it is important for us to look at whether our coordinate information is accurate to where the genus Danaus was observed and remove outliers with extremely poor accuracy (ie those with large positional accuracy values).

```{r plot positional accuracy vs lat & long, fig.width=14, fig.height=7}
par(mfrow=c(1,2))

# plot positional accuracy vs latitude
plot(butterfly_data$positional_accuracy ~butterfly_data$latitude, 
     xlab = "Latitude", 
     ylab = "Positional Accuracy", 
     pch = 19, 
     col = adjustcolor(1,0.4),
     main = "Positional Accuracy vs Latitude")

# plot positional accuracy vs longitude
plot(butterfly_data$positional_accuracy ~butterfly_data$longitude, 
     xlab = "Longitude", 
     ylab = "Positional Accuracy", 
     pch = 19, 
     col = adjustcolor(1,0.4),
     main = "Positional Accuracy vs Longitude")

```

There are some outliers in our data, in particular in the data points that contain latitude or longitude greater than 30,000 meters away from observations. We clean the data by removing data points that are more than 15,000 meters away from the recorded location.

```{r plot positional accuracy vs lat & long w/o outliers, fig.width=14, fig.height=7}
# remove outliers
butterfly_data <- butterfly_data %>% filter(positional_accuracy < 15000)

par(mfrow=c(1,2))

# plot positional accuracy vs latitude
plot(butterfly_data$positional_accuracy ~butterfly_data$latitude, 
     xlab = "Latitude", 
     ylab = "Positional Accuracy", 
     pch = 19, 
     col = adjustcolor(1,0.4),
     main = "Positional accuracy vs Latitude")

# plot positional accuracy vs longitude
plot(butterfly_data$positional_accuracy ~butterfly_data$longitude, 
     xlab = "Longitude", 
     ylab = "Positional Accuracy", 
     pch = 19, 
     col = adjustcolor(1,0.4),
     main = "Positional accuracy vs Longitude")
```
By removing the outliers, the maximum positional accuracy present in our dataset is improved, and we can be more certain that our dataset's longitude and latitude values are reasonably close to the actual observation locations.


### Species distribution across the world
#### Adding colours qualitatively with a library
```{r add colours qualitatively}
butterfly_data$scientific_name <- as.factor(butterfly_data$scientific_name)
# set up colors for butterflies
jColors <- with(butterfly_data,
                data.frame(scientific_name = levels(scientific_name),
                           color = I(brewer.pal(nlevels(scientific_name), name = 'Accent'))))
# set up new data frame for plot
butterfly_data_scatter <- data.frame(subset(butterfly_data, select = c(latitude, longitude, scientific_name)),
           matchRetVal = match(butterfly_data$scientific_name, jColors$scientific_name))
```

#### Plot points onto world map
```{r plot points onto world map, fig.width=14, fig.height=9}
# create map
plot(getMap(), 
     xlab = "Longitude",     
     ylab = "Latitude",     
     main = 'Genus Danaus (Tiger MilkWeed Butterfly) Observed Across the World')

# add data points to map
points(butterfly_data_scatter$longitude, butterfly_data_scatter$latitude, 
       col=alpha(jColors$color[match(butterfly_data_scatter$scientific_name, jColors$scientific_name)],0.4), 
       pch=19,cex=1.3)

# add legend to map
legend("bottom", inset=c(0,0), 
       legend = as.character(jColors$scientific_name),
       col = jColors$color, pch = 19, cex=0.7,
       xpd=TRUE, horiz=T)
```
There is a concentration of observations from the USA, in particular of the species Danaus plexippus. Additionally, there is a lack of samples from the eastern continents in the world. Even with this disparity in available data, we still observe differences in the distribution of Danaus species spread across the globe. For example, we see that Danaus petilia is only observed in Australia in this data set. This species is typically found in Australia and tropical countries.

## Statistical Tests
### Chi-squared goodness of fit test for metamorphic stages
Based on the number of samples for each metamorphic stage in our data set, we decided to test the null hypothesis that there is no fit between the proportion of samples in each stage that we see and the expected metamorphic stage durations. This will tell us if our stage data is skewed. Because it is categorical data, we conducted a Chi-squared goodness of fit test.

First, we found the average duration of butterflies' metamorphic stages at: https://www.joyfulbutterfly.com/life-cycle-of-a-butterfly/

We make the assumption that the following information applies for genus Danaus across the world, time, and region.
<pre>
# butterfly life cycle

Egg:               5 days                              5/57.5 = 0.08695652

Caterpillar:       3.5 weeks = 24.5 days               24.5/57.5 = 0.426087

Chrysalis/Pupa:    1.5 weeks = 10.5 days               10.5/57.5 = 0.1826087

Adult butterfly:   2.5 weeks = 17.5 days               17.5/57.5 = 0.3043478

Total lifecycle:   5+24.5+10.5+17.5 = 57.5 days
</pre>

```{r chi-squared test}
chisq.test(table(butterfly_data$metamorphosis_stage), p = c(5, 24.5, 10.5, 17.5)/57.5)
```
Based on the chi-square goodness of fit test, our p-value is 2.2e-16, which is less than 0.005, meaning this test strongly rejects the null hypothesis of no fit. Thus, we conclude that our observed dataset for genus Danaus matches with the provided, predicted average butterfly lifecycle durations.

### Linear regression for observation month vs. latitude
In the "Metamorphic stage of sample compared to observation date" section, we suggested that the apparent bimodal distribution of observations by month may be caused by a connection between observation month and observation location. Here, we hypothesize that a sample's observation month is related to the latitude of the observation location, because butterflies are most active in warm weather.

```{r month observed vs latitude observed on}
# linear regression test
lr <- lm(butterfly_data$month~butterfly_data$latitude)
summary(lr)
```
Based on the results of the linear regression, we cannot reject the null hypothesis that here is no significant correlation between observation month and latitude, because our p-value of 0.236 is > 0.05.

## Summary of Results
The genus Danaus dataset contains interesting observations of metamorphosis stages around the world and over time. We verified that the proportion of metamorphosis stages in our data fit the expected duration of the stages using a chi-squared goodness of fit test. The resulting p-value of 2.2e-16 allowed us to strongly reject the null hypothesis of no fit.

We noticed that the observation months appear to have a bimodal distribution, with the first peak in March and the second peak in August. We hypothesized that these two distributions indicate the peak butterfly seasons in two different climates. Because butterflies are most abundant in warm weather, we tested the null hypothesis that there is no significant correlation between observation month and latitude using a linear regression. Based on the resulting p-value of 0.236, we cannot reject the null hypothesis. This does not mean that latitude has no effect on observation month, only that it does not have a significant effect on it.

## Contribution Summary
Kate Skocelas created the metamorphic stages plot and the stages by month and year plots. She also conducted the linear regression test.
Julia Zheng created the positional accuracy plots and the world map plot. She also conducted the chi-squared goodness of fit test.

